---
title: "Macroinvertebrate Data Analysis"
author: "Marney Pratt"
date: "March 27, 2023"
format: 
  html:
    code-overflow: wrap
---

# Getting Setup

Keep all of your analysis in this same .qmd file and keep all the associated files in the project folder. Make sure the project name is showing in the upper right of the RStudio window.

## Load packages

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse) ## for readr, dplyr, ggplot2
#library(vegan) ## for calculating diversity measures (you can remove this if you are not calculating diversity measures)
library(ggbeeswarm) ## for jittering points (you can remove this is you did not use a dotplotv)
library(flextable) ## for making formatted tables that can be pasted into Word or Google Docs

```

## Import Data

Figure out which data file(s) you need to import (see the metadata files for more information about each data file). Find the code you need in the "importing_data" folder within the "script_templates" folder. Copy and paste the code you need into the chunk below.

```{r}
#| label: import data

env <- read_csv("data/env.csv",
                 col_types = cols(date = col_date(format = "%Y-%m-%d"),
                  
                  location = col_factor(levels = c(
                    "Downstream","Upstream")), 
                  
                  season = col_factor(levels = c("Summer","Fall")), 
                  
                  year = col_factor()
                 )) 

macros <- read_csv("data/macros.csv",
                 col_types = cols(date = col_date(format = "%Y-%m-%d"),
                  
                  location = col_factor(levels = c(
                    "Downstream","Upstream")), 
                  
                  season = col_factor(levels = c("Summer","Fall")), 
                  
                  year = col_factor()
                 )) 


master.taxa <- read_csv("data/master.taxa.csv",
                        show_col_types = FALSE) %>% 
  distinct()


```

# Data Wrangling

Find the code you need to prepare or "wrangle" your data to make any calculations needed and combine any datasets you need to combine by looking in the "data_wrangling" folder within the "script_templates" folder. Copy and paste the code you need in the chunk below.

```{r}
#| label: data wrangling
#| message: false
#| warning: false

#calculate limpet density for each sampleID
limpet.density <- macros %>% 
  
  #join taxa info
  left_join(., master.taxa) %>%
  
  # Summarize for each sampleID and each family 
  # change group_by function to remove or add grouping variables as needed 
  # if you want a specific group of organisms, add that column name into 
  # the list of grouping variables (family, organism_aggr, FFG, etc)
  # then filter for the organism or group you want
  group_by(sampleID, family) %>% 
  dplyr::summarise (density = sum(invDens, na.rm = TRUE)) %>% 

  #fill in group combos where there were none present with zeros
  ungroup() %>% 
  complete(sampleID, family,
           fill = list(density = 0)) %>% 
  
  #filter for limpets
  filter(family == "Ancylidae")

#get sample info
sample.info <- macros %>%
  select(date, sampleID, season, year, location, benthicArea) %>% 
  distinct()

#add sample info back to density data
limpet.density <- left_join(limpet.density, sample.info)


#select desired variables from env
env2 <- env %>% 
  select(date, sampleID, season, year, location, 
         pH, DO, wTemp, light, flow, turb, cond, 
         mon.precip, mon.ADD, max.discharge) 

#add env variables to density data
limpet.df <- left_join(limpet.density, env2) %>% 
  filter(year != "2018")



```

# Data Description

## Summary Table of Sampling Effort

Make a table of the data that you used in your analysis (how many samples and how much area was sampled in each of the groups) to include in the Data Analysis Methods section of your paper. Go to the "data_description" script templates and select the correct version of the 'sample_table' script based on which factors you are using and paste that code below.

```{r}
#| label: sample table
#| tbl-cap: Unformatted Summary Table
#| message: false
#| warning: false

table.sum <- limpet.df %>%  #put the name of your wrangled data here
  
  #summarize by date, season, & location (to get number of microhabs and area sampled per date)
  group_by(date, season, location) %>% 
  summarize(
    area = sum(benthicArea),
    samples = n()
  ) %>% 
  
  #summarize by season & location 
  group_by(season, location) %>% 
  summarize(
    dates = str_c(date, collapse = ", "),
    samples = sum(samples),
    area = sum(area),
  )  


table.sum


```

If you want code to make a formatted table for you, go to the "data_description" script templates and select the correct version of the 'formatted_sample_table' script based on which factors you are using and paste that code below.

```{r}
#| label: formatted sample table
#| tbl-cap: Summary Table
#| include: true


#To get a formatted table, make sure that the {flextable} package is loaded
#Use additional code to create the summary table called "table.sum"
#If you want to format the table yourself, then don't use this code

#see https://davidgohel.github.io/flextable/ for more information and formatting options


#create the formatted table
ft <- flextable(table.sum)

#create header labels
ft <- set_header_labels(ft,
                        season = "Season",
                        location = "Location",
                        dates = "Dates",
                        samples = "# of Samples",
                        area = "Total Area Sampled (m^2)"
                        )

#format superscript
ft <- compose( 
  ft, j = "area", part="header",
  value = as_paragraph(
    "Total Area Sampled (m", 
    as_sup("2"),
    ")"
  )
)

#bold the headings
ft <- bold(ft, part="header")

ft <- merge_v(ft, j = ~ season)

#set table properties
ft <- set_table_properties(ft, layout = "autofit", width = 1)


#add lines between year x season
ft <- theme_vanilla(ft) %>% 
  fontsize(size = 10)

#center columns
ft <- align(ft, align = "center", part = "all" )

ft <- fix_border_issues(ft)

#print the table
#right click on the table, choose select all, 
#choose copy, then paste in your document
#finish formatting as needed in your document
ft

```

## Weather Table

This is optional, but if you want to have a summary of what the weather was like in the month before sampling, then use the two code chunks below along with the template code from "script_templates" then "data_description" and select the correct version of the "weather_table\_.R" script based on which factors you are using and paste that code below. (note that the weather is the same for both locations, so you do not need that factor)

```{r}
#| label: weather table
#| tbl-cap: Unformatted Weather Table
#| message: false
#| warning: false

#before you run this code, make sure your dataframe has all the necessary columns
#you need season, year, location, mon.ADD, mon.precip

weather.sum <- limpet.df %>%  #put the name of your wrangled data here
    group_by(season, year) %>% 
  summarize(
    mon.ADD = round(mean(mon.ADD, na.rm=TRUE), digits=0),
    mon.precip = round(mean(mon.precip, na.rm=TRUE), digits=0)
  )  

weather.sum

```

If you want code to make a formatted table for you, go to the "data_description" script templates and select the 'formatted_weather_table\_.R' script based on which factors you are using and paste that code below. (note that the weather is the same for both locations, so you do not need that factor)

```{r}
#| label: formatted weather table
#| tbl-cap: Weather Table
#| include: true

#To get a formatted table, make sure that the {flextable} package is loaded
#If you want to format the table yourself, then don't use this code

#see https://davidgohel.github.io/flextable/ for more information and formatting options

#create the formatted table
ft <- flextable(weather.sum)

#create header labels
ft <- set_header_labels(ft,
                        season = "Season",
                        year = "Year",
                        mon.ADD = "Monthly Accumulated Degree Days",
                        mon.precip = "Monthly Total Precipitation (mm)"
                        )

#bold the headings
ft <- bold(ft, part="header")

ft <- merge_v(ft, j = ~ season + year)

#set table properties
ft <- set_table_properties(ft, layout = "autofit", width = 1)

#add lines between year x season
ft <- theme_vanilla(ft) %>% 
  fontsize(size = 10)

#center columns
ft <- align(ft, align = "center", part = "all" )

ft <- fix_border_issues(ft)

#print the table
#right click on the table, choose select all, 
#choose copy, then paste in your document
#finish formatting as needed in your document
ft


```

## Descriptive Statistics

Use the template code below to calculate the descriptive statistics for your variable of interest. The output of this code should not all be reported in your paper. Decide which (if any) of the descriptive statistics you should report and make sure to round the values. Descriptive statistics can be reported in the text of the results (see this [example results section](https://docs.google.com/document/d/1vwyOSmhZzTuIfAEbpeiv3yCsU1UMZDsTXgnbE80nPTU/edit?usp=sharing)) if relatively simple, or you can make a formatted table of just the necessary, rounded values to include in your paper (in which case you might want to use "formatted_descriptiveStats_table.R" in the "data_description" script templates to help you make a formatted table).

```{r}
#| label: descriptive stats
#| message: false
#| warning: false

# the code below will calculate descriptive statistics for a variable of interest grouped by another variable

density.sum <- limpet.df %>% # put the name of the data frame here
  filter(!is.na(density)) %>% # remove missing values from the variable of interest
  group_by(season, location) %>% # put the name of the grouping variable(s) here
  summarize(mean = mean(density), # put the name of the variable you want to summarize in this & following blanks
            median = median(density), 
            SD = sd(density), 
            IQR = IQR(density), 
            min = min(density),
            max = max(density),
            N = n())

density.sum


DO.sum <- limpet.df %>% # put the name of the data frame here
  filter(!is.na(DO)) %>% # remove missing values from the variable of interest
  group_by(season, location) %>% # put the name of the grouping variable(s) here
  summarize(mean = mean(DO), # put the name of the variable you want to summarize in this & following blanks
            median = median(DO), 
            SD = sd(DO), 
            IQR = IQR(DO), 
            min = min(DO),
            max = max(DO),
            N = n())
DO.sum


wTemp.sum <- limpet.df %>% # put the name of the data frame here
  filter(!is.na(wTemp)) %>% # remove missing values from the variable of interest
  group_by(season, location) %>% # put the name of the grouping variable(s) here
  summarize(mean = mean(wTemp), # put the name of the variable you want to summarize in this & following blanks
            median = median(wTemp), 
            SD = sd(wTemp), 
            IQR = IQR(wTemp), 
            min = min(wTemp),
            max = max(wTemp),
            N = n())
wTemp.sum

```

# Data Visualization

Find the code you need to visualize (=graph) your data by looking in the "graphing" folder within the "script_templates" folder. Copy and paste the code you need in the chunk below.

```{r}
#| label: limpet density vs DO graph
#| message: false


# this code will make a scatter plot with each graphed on a log-scale axis

limpet.df %>% 
  filter(!is.na(DO)) %>% 

ggplot(
  data = ., 
  aes(x = DO, y = density+1, color = location)) + 
  geom_point(size = 3) +                   #play with the point attributes as needed
  geom_smooth(method = "lm", se = FALSE) + #adds line of best fit
  ylab(bquote("Density (number/"~m^2*")")) +
  xlab("Dissolved Oxygen (mg/L)") +
  labs(color = "Location") +
  theme_classic(base_size = 18) +
  scale_y_log10() +     #changes y-axis to log-scale
  annotation_logticks(sides = "l") + # adds log-scale tick marks 
  facet_wrap(vars(season))   

#see the resources below for some additional options to make a nice scatter plot 
# https://www.datanovia.com/en/lessons/ggplot-scatter-plot/
# https://ggplot2.tidyverse.org/index.html

```

To save your graph with the `ggsave()` function, you need to name the resulting file with surrounding " ", and indicate the size via height, width, and units. Don't forget to save the graph with a dpi call between 300-500 to make it nice and crisp! Look at the `ggsave()` help file for more information and options.

```{r}
#| eval: false


# save the graph!
ggsave(filename="images/limpet.DO.png",  #recommended to use.png or .jpg file types
       height = 5, width = 8, units = "in", 
       dpi = 300)

```

```{r}
#| label: limpet density vs wTemp graph
#| message: false


# this code will make a scatter plot with each graphed on a log-scale axis

limpet.df %>% 
  filter(!is.na(wTemp)) %>% 

ggplot(
  data = ., 
  aes(x = wTemp, y = density+1, color = location)) + 
  geom_point(size = 3) +                   #play with the point attributes as needed
  geom_smooth(method = "lm", se = FALSE) + #adds line of best fit
  ylab(bquote("Density (number/"~m^2*")")) +
  xlab("Water Temperature ("~degree~"C)") +
  labs(color = "Location") +
  theme_classic(base_size = 18) +
  scale_y_log10() +     #changes y-axis to log-scale
  annotation_logticks(sides = "l") + # adds log-scale tick marks 
  facet_wrap(vars(season))   

#see the resources below for some additional options to make a nice scatter plot 
# https://www.datanovia.com/en/lessons/ggplot-scatter-plot/
# https://ggplot2.tidyverse.org/index.html
```

```{r}
#| eval: false


# save the graph!
ggsave(filename="images/limpet.wTemp.png",  #recommended to use.png or .jpg file types
       height = 5, width = 8, units = "in", 
       dpi = 300)

```

```{r}

#| label: wTemp vs season and location

limpet.df %>% 
  filter(!is.na(wTemp)) %>% 

ggplot(
  data = .,   
  aes(x = location, y = wTemp,   
      fill = location, color = location)) + 
  geom_quasirandom( 
    shape=21, size=2, alpha = 0.5, width=0.2, 
    dodge.width = 1,  show.legend = FALSE) + 
  stat_summary(fun = median, 
               fun.min = median, 
               fun.max = median, 
               geom = "crossbar", na.rm = TRUE,
               width = 0.5, size = 0.5, 
               color = "black",
               position=position_dodge(width=1), 
               show.legend = FALSE) + 
  ylab("Water Temperature ("~degree~"C)") +
  xlab("Location") +
  coord_cartesian(expand=TRUE) +
  theme_classic(base_size=16) +
  facet_wrap(vars(season))   

```

```{r}
#| eval: false

# save the graph!
ggsave(filename="images/wTemp.season.png",  #recommended to use.png or .jpg file types
       height = 5, width = 8, units = "in", 
       dpi = 300)

```

```{r}
#| label: DO vs wTemp graph
#| message: false

limpet.df %>% 
  filter(!is.na(DO), !is.na(wTemp)) %>% 

ggplot(
  data = ., 
  aes(x = wTemp, y = DO, color = location)) + 
  geom_point(size = 3) +                   #play with the point attributes as needed
  geom_smooth(method = "lm", se = FALSE) + #adds line of best fit
  ylab("Dissolved Oxygen (mg/L)") +
  xlab("Water Temperature ("~degree~"C)") +
  labs(color = "Location") +
  theme_classic(base_size = 18) 


```

```{r}
#| eval: false

# save the graph!
ggsave(filename="images/wTemp.DO.png",  #recommended to use.png or .jpg file types
       height = 5, width = 8, units = "in", 
       dpi = 300)

```

# Citing R in Your Paper

In the Data Analysis Methods section of your paper, you need to cite how the data were analyzed including (1) any calculations done (for example, how was density calculated?), and (2) what software was used for the analysis.

Here is a detailed explanation of How to Cite R in your Methods.

You need to cite base R but how will you know what version of R you are using?

You should also cite the **most important packages used**.

All of you should cite the {tidyverse} super package since we used {reader}, {dplyr}, and {ggplot2} to import, wrangle, and graph our data. If you used the dot plot in your paper, then you also used {ggbeeswarm} which is NOT part of the {tidyverse} so it needs to be cited separately. If you used the formatted table in your paper, then cite the {flextable} package. If you used {vegan} to calculate any diversity indices then make sure to cite it.

To properly cite packages, you need to find out the information needed for each package. Run the code below to determine the version of R and packages used. (note that you do NOT need to cite ALL of the packages that the `sessionInfo` function spits out. See above for what you need to cite)

```{r}
#| label: sessionInfo

sessionInfo()

```

The above is the information you need to cite which version of R as well as any packages you used. While the version number is necessary, you also need more information to cite R and the packages. Refer to [How to Cite R in your Methods](https://docs.google.com/presentation/d/1RG4_R-MDOy1vbMz-M1gfOcMtgqWu9Kvv_R4ex5QdcVk/edit?usp=sharing) to help you figure out how to cite in the text as well as the full citation in the Literature Cited section at the end of your paper.

Use the code below to get more necessary information to cite R itself

```{r}
#| label: cite R

#citation info for base R 
citation()

```

Use the code below to cite the {tidyverse} package and then repeat the code replacing "tidyverse" with any other packages you need to cite

```{r}
#| label: cite tidyverse

#citation info for base R 
citation("tidyverse")

```

## Rendering

Don't forget to "render" your .qmd file when you are done. Make sure that each code chunk has an unique label name or no name before you render. Make sure all your files are in the project folder.

Render straight to HTML, open the HTML file in your browser, and then "print" to a PDF file.

Insert a file printout of the PDF in the "Project Data Analysis" section of your Lab Notebook.
